TrenchR: an R package for transparent environmental and ecological biophysics

### 2020-11-05 Manuscript_full.Rmd
update text

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi"
date: "February 4, 2020"
output:
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

@@

Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground. However, climate change responses often hinge on the occurrence of acute thermal stress events due to the radiative heating of organisms near the ground. We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR). The package facilitates microclimate modeling to translate weather station data into the environmental conditions experienced by organisms and biophysical modeling to predict organismal body temperatures given the environmental conditions. The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules introducing microclimate and biophysical modeling.

## Introduction
Many responses of species and ecosystems to climate change are inconsistent with predictive models [@maguire2015modeling]. Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic]. Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m). However, air temperatures near the ground where most organisms reside can be considerably warmer (XX-XXC) and absorption of solar radiation can raise body temperatures well above air temperatures [XXXX].  Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions. Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life]. We aim to help remedy these shortcomings by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of the tools has not kept pace with research on climate change responses. Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].  Increasing the dissemination of computational tools is facilitating adoption. The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models developed by Warren Porter and colleagues over decades. Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.  The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.  Although many of the functions are general, some components of TrenchR are best suited for ectothermic animals. We have minimized the inclusion of plant-specific biophysical models since they are the focus of several R packages [tealeaves: https://www.biorxiv.org/content/biorxiv/early/2019/01/24/529487.full.pdf; plantecophys [@duursma2015plantecophys]].

## Components
The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users. Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github. Issues and feature requests can be contributed in Github [XXXX]. The package is maintained and checked using Travis CI as a tool for continuous integration. TrenchR functions are organized into the following categories:

**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.

@@

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss. Functions serve to aggregate these forms of heat exchange into energy balances and to use the energy balances to predict operative environmental temperatures. Operative environmental temperatures are the predicted temperatures of organisms with specified physical properties based on the above components of heat exchange [XXXX]. [WATER BALANCE.]

**Specific biophysical functions** provide biophysical models that have been built for specific organisms based on their physical properties, behavior, environment, and life history. We currently provide published biophysical models for lizards, butterflies, and grasshoppers and plan to include models for marine taxa (mussels and limpets). 
APIs??

We do not present each function’s biophysical equations for tractability, but most functions include a reference deriving the equations. We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available. Many functions additionally include comments that describe potential parameterizations. Biophysical texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR.
[Table 8.2]

## Tutorial and introductory modules
We introduce each function in categorically grouped tutorials.  A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* tutorial provides resources for estimating the environmental conditions experienced by organisms.  This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles. Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*. Components of an energy budget can be estimated using individual functions and then operative environmental temperatures, $T_e$, can be solved for using either a generic energy balance or taxa specific biophysical models.

For a broader background on microclimate and biophysical modeling, we have updated a series of modules focused on “Physical processes in ecosystems”.  We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical]. The modules provide less detailed coverage of many of the topics contained in the books. The series of 14 modules starts with an overview of the calculus and physics principles underlying the modeling. Thermodynamics and energy budget modeling are then introduced with detailed examples. We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R codes and utilization of TrenchR functions. We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [-@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad-scale climatic limits. The modules are provided in HTML form at a bookdown site and as a pdf. The R markdown files are available in Github at https://github.com/trenchproject/PhysicalProcessesEcosystems/.

## Implementation
We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial. We first estimate the energy budget with component functions and we then use an integrated biophysical model to estimate operative environmental temperatures. We start by generating environmental inputs.

We will use the energy budget to estimate body temperatures, $T_b$, of organisms can depart dramatically from the air temperatures due to energy exchange with the environment.  Heat is gained from absorbing solar and thermal radiation and from metabolic reactions. Heat is lost through the organism's emission of radiation and the evaporation of water. The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. The balance of these heat exchanges can be estimated and is referred to as operative environmental temperature, $T_e$. $T_e$ indicates the steady-state temperature of an organism with given physical properties in a particular environment [@bakken1992measurement].

Let's assume the lizard is in an unshaded location where the daily air temperature varies from a minimum of 10°C to a maximum of 25°C, the soil surface temperature varies from a minimum of 15°C to a maximum of 30°C, and the windspeed is 0.5 m/s.

Assume that atmospheric transmissivity $\tau=0.7$, albedo $\rho=0.6$

We assume $T_b=T_a$ to illustrate calculations

@@

#set up input data as variables
lat=35.69; lon= -105.944; elev=2121
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30
V=0.5
#assumptions
tau=0.7; rho=0.6
Tb0=25

doy= day_of_year("2018-06-01", format= "%Y-%m-%d")
snoon= solar_noon(lon=lon, doy=doy)
dayl= daylength(lat=lat, doy=doy)
tr= snoon-dayl/2
ts= snoon+dayl/2
```

We can estimate hourly solar radiation as follows:
```{r}
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon)
psi_rad= degree_to_radian(psi_deg) #convert to radians
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]
Sdif= Srad[2,]
Sref= Srad[3,]
```

@@

### Energy balance
We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} + Q_{conv} + Q_{cond} + Q_{met} - Q_{evap},$$
We will estimate each term on the right side of the equation in turn. Estimating $Q_{abs}$requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical].

>>>>

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi,..."
date: "February 4, 2020"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

@@

Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground. However, most organisms live near the ground and acute exposure to solar radiation can substantially elevate their body temperatures. We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR). The package includes microclimate models to vertically scale weather station data to organismal heights. The biophysical modelling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa. Additional functions model and temporally partition air and soil temperatures and solar radiation. The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules.

## Background
Many responses of species and ecosystems to climate change are inconsistent with predictive most current models [@maguire2015modeling]. Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic]. Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m). However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [Kearney et al., 2009; Sinclair et al., 2016; Sunday et al., 2014].  Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions. Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life]. We aim to help remedy these shortcomings by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of the tools has not kept pace with research on climate change responses. Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].  Increasing dissemination of computational tools is facilitating adoption. The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models. Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.  The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.  Some energy budget calculation components of biophysical models are available in the ThermImage package (Tattersall, 2017). Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals. We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: (Muir, 2019); plantecophys [@duursma2015plantecophys]].

## Methods and features
The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users. Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github. Issues and feature requests can be contributed in Github (https://github.com/trenchproject/TrenchR/). The package is maintained and checked using Travis CI as a tool for continuous integration.

Functions were adapted from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles. The source reference is included in each function’s header. We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available. Many functions additionally include comments that describe potential parameterizations. The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR [Table 8.2XXXX].

### Components
TrenchR functions are organized into the following categories:
**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.

@@

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss. Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.

**Specific biophysical functions** provide biophysical models that have been built for specific organisms based on their physical properties, behavior, environment, and life history. We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails. Most models predict operative environmental temperatures, which are the equilibrium body temperatures of organisms with specified physical properties in a specific microenvironment and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].

### Vignettes and introductory modules
We introduce the function in several vignettes.  A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* tutorial provides resources for estimating the environmental conditions experienced by organisms.  This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles. Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*. Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we have updated a series of educational modules entitled “Physical processes in ecosystems”.  We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical]. The modules provide less detailed coverage of many of the topics contained in the books. The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling. Thermodynamics and energy budget modelling are then introduced with detailed examples. Modules address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. We have revised the modules (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits. The modules are provided in html form and available for download as a pdf at the bookdown server (https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/). The R markdown files are available in Github at https://github.com/trenchproject/TrenchRmodules.

## Examples
We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial. We first estimate the energy budget with component functions and we then use an integrated biophysical model to estimate operative environmental temperatures. We start by generating environmental inputs.

We will use the energy budget to estimate the lizard's body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment.  Heat is gained from absorbing solar and thermal radiation and from metabolic reactions. Heat is lost through the organism's emission of radiation and the evaporation of water. The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. The balance of these heat exchanges can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].

Let's assume the lizard is in an unshaded location where the daily air temperature varies from a minimum of 10°C to a maximum of 25°C, the soil surface temperature varies from a minimum of 15°C to a maximum of 30°C, and the windspeed is 0.5 m/s. We assume that atmospheric transmissivity $\tau=0.7$ and albedo $\rho=0.6$

@@

#set up input data as variables
lat=35.69; lon= -105.944; elev=2121 #latitude and longitude (degrees) and elevation (m)
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 #minimum and maximum of air and soil temperatures, respectively (C)
V=0.5 #wind speed (m/s)
#assumptions
tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
Tb0=25 #initial assumption of body temperature (C)

doy= day_of_year("2018-06-01", format= "%Y-%m-%d") 
snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
dayl= daylength(lat=lat, doy=doy) #estimate daylength
tr= snoon-dayl/2 #time of sunrise
ts= snoon+dayl/2 #time of sunset
```

We can estimate hourly solar radiation as follows:
```{r}
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) #estimate zenith angle (degrees)
psi_rad= degree_to_radian(psi_deg) #convert to radians
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho) #estimate radiation
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]
Sdif= Srad[2,]
Sref= Srad[3,]
```

@@

### Energy balance
We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} + Q_{conv} + Q_{cond} + Q_{met} - Q_{evap},$$
We will estimate each term on the right side of the equation in turn. Estimating $Q_{abs}$requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical]. We will initially assume $T_b=T_a$ to illustrate calculations.
